---
#title: "calibrateR"
#author: "Greg Bedwell"
#date: "`r format(Sys.time(), '%d %B, %Y')`"
permalink: /tools/calibrateR/
output:
  html_document
#  md_document:
#    variant: gfm
#    preserve_yaml: true
tags:
  - standardization
  - R
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(base.dir = "/Users/gbedwell/Documents/github/gbedwell.github.io", base.url = "/")
#knitr::opts_chunk$set(fig.path = "images/")
```

### Description
calibrateR is a package written to streamline common laboratory calculations. At the moment, this includes SEC calibration, BCA calibration, protein and nucleic acid concentration determination using UV/Vis spectroscopy, and the estimation of protein parameters given just a primary amino acid sequence.

### Example files
More in-depth example usage of calibrateR that goes through more of the function options is provided on [GitHub](https://github.com/gbedwell/example_files).

### Installation
```{r, eval=FALSE}
devtools::install_github("gbedwell/calibrateR")
```

```{r, include=FALSE}
require(calibrateR)
```

### SEC calibration
There are two primary outputs for the <code>sec()</code> function: 

1. A calibration output that calculates a standard curve given the necessary calibrant information.

2. An analytical output that estimates the molecular weight and hydrodynamic radius of a given analyte.

Both outputs can be generated in either tabular or graphical forms.



#### Generating a calibration curve
At least 3 different pieces of information are required for calibrateR to generate a calibration curve for a size-exclusion column:

1. The position of the column void volume. This is defined as <code>void</code> in the <code>sec</code> function. This can be empirically defined by running the polymer blue dextran over the column.

2. The column volume. This is defined as <code>cv</code> in the <code>sec</code> function. This can be empirically defined by running something like 0.2% (v/v) acetone over the column.

3. The masses (<code>masses</code>) and elution volumes (<code>stds</code>) of a series of standard proteins. If using the commonly used BioRad gel filtration standards, the <code>masses</code> flag can be left <code>NULL</code>.

Toggle between tabular and graphical outputs with the <code>plot</code> flag. Setting <code>plot = "yes"</code> will generate graphical output. <code>plot = "no"</code> generates tabular output. Tabular output includes linear regression information.

To generate the calibration curve, set <code>type = "cal"</code>.

The hydrodynamic radii of the standard proteins are estimated based on their masses using the relationship described by Uversky (Biochemistry. 1993. 32. 13288-13298).


```{r, eval=FALSE}
sec(masses = c(669000, 158000, 44000, 17000, 1350), 
    void = 1.87, 
    cv = 4.29, 
    stds = c(2.08,2.58,2.98,3.30,4.06), 
    unk = NULL, 
    plot = "no",
    type = "cal")
```

```{r, echo=FALSE}
knitr::kable(sec(masses = c(669000, 158000, 44000, 17000, 1350),
                 void = 1.87, 
                 cv = 4.29, 
                 stds = c(2.08,2.58,2.98,3.30,4.06), 
                 unk = NULL, 
                 plot = "no",
                 type = "cal"), 
             align = "cccccccccc", format='html', escape=TRUE) %>%
   kableExtra::kable_styling(full_width=FALSE, position="center", font_size = 14) %>% 
   kableExtra::row_spec(0 , align="c")
```



#### Estimating analyte properties
Setting <code>type = "pred"</code> will predict the molecular weights and hydrodynamic radii of unknown samples. This is based on the calculated standard curve and analyte elution volumes.

Define the analyte elution volumes as <code>unk</code> in the function.

In the graphical output, analyte parameters are reported in labels on the plot itself.

```{r, fig.height=5, fig.width=6, fig.align='center'}
sec(masses = c(669000, 158000, 44000, 17000, 1350), 
    void = 1.87, 
    cv = 4.29, 
    stds = c(2.08,2.58,2.98,3.30,4.06), 
    unk = c(2.11, 2.49, 3.28, 3.87), 
    plot = "yes",
    type = "pred")
```


### BCA calibration
The function for BCA calibration (<code>bca</code>) is generally similar the <code>sec</code> function.

The function minimally requires:

1. The known concentrations and corresponding absorbance values for a standard dilution series (<code>prot.conc</code> and <code>std.abs</code>).

2. The absorbance value of the buffer-only control (<code>blank.abs</code>).

```{r, eval=FALSE}
bca(prot.conc = c(2, 1.5, 1, 0.75, 0.5, 0.25, 0.125, 0.025),
    blank.abs = 0.1813,
    std.abs = c(2.4644, 1.989, 1.4295, 1.1075, 0.8124, 0.499, 0.3655, 0.2315),
    abs = NULL,
    dil.fac = NULL,
    plot = "no",
    type = "cal")
```

```{r, echo=FALSE}
knitr::kable(bca(prot.conc = c(2, 1.5, 1, 0.75, 0.5, 0.25, 0.125, 0.025),
                 blank.abs = 0.1813,
                 std.abs = c(2.4644, 1.989, 1.4295, 1.1075, 0.8124, 0.499, 0.3655, 0.2315),
                 abs = NULL,
                 dil.fac = NULL,
                 plot = "no",
                 type = "cal"), 
             align = "cccccccccc", format='html', escape=TRUE) %>%
   kableExtra::kable_styling(full_width=FALSE, position="center", font_size = 14) %>% 
   kableExtra::row_spec(0 , align="c")
```

To estimate the protein concentration in unknown samples, the absorbance values and dilution factor the sample(s) must also be defined (<code>abs</code> and <code>dil.fac</code>).

The <code>plot</code> and <code>type</code> flags operated identically to the <code>sec</code> function.

```{r, fig.height=4.5, fig.width=5.5, fig.align='center'}
bca(prot.conc = c(2, 1.5, 1, 0.75, 0.5, 0.25, 0.125, 0.025),
    blank.abs = 0.1813,
    std.abs = c(2.4644, 1.989, 1.4295, 1.1075, 0.8124, 0.499, 0.3655, 0.2315),
    abs = c(2.0184, 1.1641, 0.5831),
    dil.fac = c(5, 10, 20),
    plot = "yes",
    type = "pred")
```


### Protein parameter estimation
Protein parameter estimation from the primary amino acid sequence is a useful tool for any protein chemist. There are numerous webtools that estimate various parameters in this way. The <code>prot.param()</code> function in calibrateR calculates many standard parameters as well as some less common parameters that are useful for hydrodynamic analyses.

To estimate these parameters, first define the primary amino acid sequence of the protein-of-interest.

Next, define the temperature at which the experiments are being performed. This is important for the estimation of the protein's partial specific volume (vbar) and associated parameters (e.g. the radius of a sphere with equal mass and density).

```{r, eval=FALSE}
protein <- c("MSKGEELFTGVVPILVELDGDVNGHKFSVRGEGEGDATNGKLTLKFICTTGKLPVPWPTL
VTTLTYGVQCFSRYPDHMKRHDFFKSAMPEGYVQERTISFKDDGTYKTRAEVKFEGDTLV
NRIELKGIDFKEDGNILGHKLEYNFNSHNVYITADKQKNGIKANFKIRHNVEDGSVQLAD
HYQQNTPIGDGPVLLPDNHYLSTQSVLSKDPNEKRDHMVLLEFVTAAGITHGMDELYK")

calibrateR::prot.param(protein, temp=25)
```

```{r, echo=FALSE}
protein <- c("MSKGEELFTGVVPILVELDGDVNGHKFSVRGEGEGDATNGKLTLKFICTTGKLPVPWPTL
VTTLTYGVQCFSRYPDHMKRHDFFKSAMPEGYVQERTISFKDDGTYKTRAEVKFEGDTLV
NRIELKGIDFKEDGNILGHKLEYNFNSHNVYITADKQKNGIKANFKIRHNVEDGSVQLAD
HYQQNTPIGDGPVLLPDNHYLSTQSVLSKDPNEKRDHMVLLEFVTAAGITHGMDELYK")

knitr::kable(calibrateR::prot.param(protein, temp=25), align = "cccccccccc", format='html', escape=TRUE) %>%
   kableExtra::kable_styling(full_width=FALSE, position="center", font_size = 14) %>% 
   kableExtra::row_spec(0 , align="c")
```

```
















