---
title: "rigrag"
author: "Greg Bedwell"
permalink: /tools/rigrag/
output:
  html_document
#  md_document:
#    variant: gfm
#    preserve_yaml: true
tags:
  - integration targeting
  - R
---

rigrag is an R package for identifying genes preferentially targeted or avoided for integration during HIV-1 infection. The integration frequency into each gene is compared to the expected integration frequency according to phenomenological models of random integration. Genes with integration frequencies higher than random are called upper outliers or recurrent integration genes (RIGs) and genes with integration frequencies lower than random are called lower outliers or recurrent avoided genes (RAGs). Genes with integration frequencies indistinguishable from random are called non-outliers. The package includes functions to easily import genic integration site bed files <i>en masse</i>, to easily generate matched random models of HIV-1 integration for datasets, to classify genes into their respective outlier category, to easily and reproducibly generate plots of the data, and an assortment of other functions useful for comparing samples to each other. rigrag is described in more detail in a manuscript currently under review.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::opts_knit$set(base.dir = "/Users/gbedwell/Documents/github/gbedwell.github.io", base.url = "/")
#knitr::opts_chunk$set(fig.path = "images/")
```

## Example files
Example files suitable for use with rigrag are provided on [GitHub](https://github.com/gbedwell/example_files).

These files include:

+ The gene annotation file used to match total integration sites to genes.

+ An example bash script for performing the intersection with BEDtools in a way that will output a file suitable for import into rigrag.

+ 2 total integration site bed files from HEK293T cells.

+ 2 genic integration site bed files in a rigrag-compatible format derived from the total bed files and the gene annotation file.

The following examples were created using these genic integration site bed files.

## Installation
```{r, eval=FALSE}
devtools::install_github("gbedwell/rigrag")
```

```{r, include=FALSE}
require(rigrag)
```

## Import genic integration site coordinates
### Define the file-specific parameters
1. Generate a list of file names to be imported.
2. Define a list of sample identifiers for each file. These must be listed in the same order as the file names.
  
```{r}
files <- generate_file_list("~/Documents/github/example_files/rigrag_files/example_bed/",
                            pattern="*genic.bed")

sample.ids <- list("Sample #1", "Sample #2")
```


### Import the listed files
The flag <code>collpased = TRUE</code> counts the number of integration sites in each gene in each sample and returns a collapsed dataframe containing all unique genes, the number of integration events per gene, and the fraction of genic integration in each gene. To simply view all integration sites, set the flag to <code>FALSE</code>.

```{r}
sample.df <- int_file_import(files=files, sample.ids=sample.ids, collapsed = TRUE)
```


## Calculate the random model parameters according to the matched fragmentation method.
The models of random integration used in the rigrag pipeline are phenomenological models derived from large datasets of theoretical random integration site generated <i>in silico</i>. These datasets were generated by randomly fragmenting the human genome according to either random fragmentation (e.g. acoustic shearing) or pattern fragmentation (e.g. restriction enzyme digestion). The two restriction enzyme cocktails used, MseI/BglII and NheI/AvrII/SpeI/BamHI, were chosen to match previously published integration site datasets. To generate the appropriate model, the user must define which fragmentation method to use. For random fragmentation, use the function <code>random_frag_model()</code>. For enzyme digestion, use the function <code>pattern_frag_model()</code> and define the particular cocktail with the flag <code>pattern="MB"</code> or <code>pattern="NASB"</code>.

```{r, results='hide'}
models.df <- pattern_frag_model(df=sample.df, pattern="MB")
```

The output contains the same information as the input dataframe, but with additional columns denoting the expected fraction of genic integration for an equally sized random dataset (expected.frac), the expected upper limit of integration frequency (upper.limit), the expected lower limit of integration frequency (lower.limit), and whether or not the gene scores as an upper-, lower-, or non-outlier in the given sample (outlier.type).

An brief example table is shown below:

```{r, echo=FALSE, eval=TRUE}
dat1 <- models.df %>%
  dplyr::filter(outlier.type=="UPPER")

dat2 <- models.df %>%
   dplyr::filter(outlier.type=="LOWER")

dat3 <- models.df %>%
   dplyr::filter(outlier.type=="NONE") %>%
   dplyr::sample_n(1000, replace=FALSE)
 
models.trim <- rbind(dat1, dat2, dat3) %>%
   dplyr::sample_n(100, replace = FALSE) %>%
   dplyr::mutate(across(c(frac.int, expected.frac, upper.limit, lower.limit), round, digits=8)) %>%
   dplyr::arrange(sample, gene.name)


rm(dat1, dat2, dat3)


DT::datatable(models.trim,
              fillContainer = TRUE, 
              extensions = 'FixedColumns',
              options = list(scrollX = TRUE, scrollY = "200px",
                             scrollCollapse = TRUE, autoWidth = FALSE))
```

## Output model plots
The <code>model_plot()</code> function in rigrag is intended to quickly output a visualization of the data contained within the <code>*_frag_model()</code> output. The generated plots depict 4 important things:

1. The calculated random model. This is depicted as the gray shaded region. The expected integration frequency is shown as the straight line near the middle of the shaded region. The upper and lower boundaries of the model are depicted above and below the expected integration frequency, respectively.

2. Upper outliers (RIGs). These are depicted as blue data points. These represent the genes integrated into <i>more</i> frequently than expected.

3. Lower outliers (RAGs). These are depicted as red data points. These represent the genes integration into <i>less</i> frequently than expected

4. Non-outliers. These are depicted as gray data points. These represent genes integrated into at frequencies indistinguishable from random. 

```{r, fig.align = "center", fig.width=12, fig.height=5.5, fig.retina=2, warning=FALSE}
model_plot(df=models.df,
           output.type="combo", 
           num.col=2, 
           labels="AUTO", 
           label.size=17,
           write.out = "no")
```








