---
title: 'The ECF, the 3 point shot, and empirical Bayes'
excerpt: 'Analyzing 3 point shooting percentage in the 2023 ECF.'
classes: wide
output:
  md_document:
    variant: gfm
    preserve_yaml: true
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = paste0(
        '/Users/gbedwell/Documents/github/gbedwell.github.io/_posts/', '2023-06-12', '-', 'ecf-three','.md'
      ),
      envir = globalenv()
    )
  })
---

```{r, include=FALSE}
date <- "2023-06-12"

base.dir <- "/Users/gbedwell/Documents/github/gbedwell.github.io/"
base.url <- "/"
dir.create(paste0("../figures/", date))
fig.path <- paste0("../figures/", date, "/")

knitr::opts_chunk$set(
  fig.asp = 5/7,
  fig.width = 7,
  dpi = 300,
  fig.align = "center",
  out.width = "80%",
  comment = "#>",
  collapse = TRUE,
  dev = "png",
  base.dir = base.dir, 
  base.url = base.url,
  fig.path = fig.path
  )
```

### Introduction
The 2022-2023 NBA playoffs have come to an end. The Denver Nuggets, this season's Western Conference champions and Finals favorites, have defeated the never-say-die Miami Heat 4 games to 1. The Nuggets, led by the virtually unstoppable Nikola Jokíc (who <i>averaged</i> 30-13.5-9.5 per game across these playoffs), made relatively easy work of the 8-seed Heat. Nevertheless, the Heat's impressive, and at times utterly dominant, run through both the 1-seed Milwaukee Bucks (4-1) and the 2-seed Boston Celtics (4-3) has already had [significant consequences](https://www.nba.com/bucks/news/milwaukee-bucks-part-ways-with-head-coach-mike-budenholzer) and could yet have more.

In the Eastern Conference Finals (ECF) against the Celtics, the Heat overturned double-digit deficits in the second halves of both Games 1 and 2 and went on to win them both. Game 3 was an emphatic 128-102 Heat victory. The Celtics won the next 3 games, evening the series at three games each and forcing a deciding Game 7 back in Boston. The Heat absolutely embarrassed the home team and silenced the TD Garden crowd with a 103-84 victory. Much has been said about how this ECF played out. Most of this commentary can be succinctly summarized as the Heat having "that dawg in 'em", while the Celtics, apparently, do not. Another, more tangible, aspect of the series, however, was the apparent difference in 3 point shooting percentage between the Heat and the Celtics. Per [basketball-reference](https://www.basketball-reference.com/), the Celtics rode the three point shot to 57 wins over the 2022-2023 NBA season, shooting 42.6 threes per game (2nd overall) with a season average 37.7% success rate (6th overall). The Heat, on the other hand, shot 34.8 threes per game (10th overall) with a measly 34.4% success rate (27th overall). During the ECF, however, the Heat shot 43.4% from behind the arc, while the Celtics shot just 30.3%. This certainly seems like a staggering reversal. Given the comparatively small sample size of the ECF, however, it's difficult to compare these numbers directly to the 2022-2023 season averages. 

I don't want to make this a post about empirical Bayes, so I will leave a lot of the nitty gritty details out of the published version. You can, however, find the entire .Rmd file [here](https://github.com/gbedwell/gbedwell.github.io/blob/master/_R/ecf-three.Rmd).

All of the data used in this post was retrieved from the fantastic [basketball-reference.com](https://www.basketball-reference.com/). 

```{r, echo = FALSE}
# Load packages
pkgs <- c( "gamlss", "rvest", "dplyr", "tibble", "tidyr", "ggplot2", "broom.mixed")

invisible( 
  sapply( X = pkgs,
          FUN = function(x){
            suppressPackageStartupMessages(
              library( x, quietly = TRUE, character.only = TRUE )
              )
            }
          )
  )
```

```{r, echo = FALSE}
# Get regular season data for the past 5 seasons.

season <- lapply( X = as.list( c( 2019, paste0( "202", 0:3 ) ) ),
                  FUN = function(x){
                    url <- paste0( "https://www.basketball-reference.com/leagues/NBA_", 
                                   x, "_totals.html" )
                    page <- rvest::read_html( url )
                    
                    cols <- page %>% 
                      rvest::html_nodes("table#totals_stats > thead > tr > th") %>%
                      rvest::html_text()
                    
                    cols <- cols[ -c(1) ]
                    
                    dat <- data.frame( 
                      t( matrix( data = page %>% 
                                   rvest::html_nodes("table#totals_stats > tbody > tr > td") %>%
                                   rvest::html_text(),
                                 nrow = 29 )
                         ) 
                      )
                    colnames( dat ) <- cols
                    
                    dat <- dat |>
                      dplyr::mutate( dplyr::across( c( 3, 5:29 ), as.numeric ) ) |>
                      tibble::add_column( Season = as.numeric( x ), .after = 3 ) |>
                      dplyr::filter( Tm != "TOT" )
                    }
                  )

season <- do.call( rbind, season )
```

```{r, echo = FALSE}
# Get three point data
season.three <- season |>
  dplyr::select( Player, Season, Tm, `3P`, `3PA`, `3P%` ) |>
  tidyr::drop_na()

# Collate team-wide data
team.three <- season.three |>
  dplyr::group_by( Season, Tm ) |>
  dplyr::summarise( `3P` = sum( `3P` ),
                    `3PA` = sum( `3PA` ),
                    .groups = "drop" ) |>
  dplyr::mutate( `3P%` = `3P` / `3PA` )

# Filter for Boston and Miami 3P data.
bm.three <- team.three |>
  dplyr::filter( Season == 2023, Tm %in% c( "BOS", "MIA" ) ) |>
  dplyr::mutate( time = "Season" )
```

```{r, echo = FALSE}
# Get 2023 ECF data
ecf.url <- "https://www.basketball-reference.com/playoffs/2023-nba-eastern-conference-finals-heat-vs-celtics.html"
ecf.page <- read_html( ecf.url )

ecf <- lapply( X = as.list( c( "MIA", "BOS") ),
               FUN = function(x){
                 cols <- ecf.page %>%
                   rvest::html_nodes( paste0( "table#", x, " > thead > tr > th") ) %>%
                   rvest::html_text()
                 cols <- cols[ -c( 1:5 ) ]
                
                 dat <- data.frame( 
                   t( matrix( data = ecf.page %>% 
                                rvest::html_nodes( paste0( "table#", x, " > tbody > tr > td" ) ) %>%
                                rvest::html_text(),
                              nrow = 29 )
                      )
                   )
                 colnames( dat ) <- cols
                
                 dat <- dat[ , 1:20 ]
                
                 dat <- dat |>
                   dplyr::mutate( dplyr::across( 2:20, as.numeric ) ) |>
                   tibble::add_column( Season = 2023, .after = 1 ) |>
                   tibble::add_column( Tm = x, .after = 2 )
                 }
               )
```

```{r, echo = FALSE}
# Get ECF 3P data
ecf.three <- do.call( rbind, ecf ) |>
  dplyr::select( Player, Season, Tm, `3P`, `3PA` ) |>
  dplyr::mutate( `3P%` = `3P` / `3PA` ) |>
  tidyr::drop_na()

# Get Boston and Miami ECF 3P data
bm.ecf.three <- ecf.three |>
  dplyr::group_by( Season, Tm ) |>
  dplyr::summarise( `3P` = sum( `3P` ),
                    `3PA` = sum( `3PA` ),
                    .groups = "drop" ) |>
  dplyr::mutate( `3P%` = `3P` / `3PA`,
                 time = "Playoffs" )

# Make a 2023 Boston and Miami 3P df
bm.stats <- rbind( bm.three, bm.ecf.three )
```

```{r, echo = FALSE, eval = FALSE}
# Check for systematic trends in team-wide 3P data.
# Most obvious potential confounder is 3P attempts -- better shooting teams might take more 3s?

ggplot( team.three, aes( x = `3PA`, y = `3P%` ) ) +
  geom_hline( yintercept = mean( team.three$`3P%` ), linewidth = 1, color = "red" ) + 
  geom_point( size = 3, shape = 21, fill = "gray75", color = "black" ) +
  theme_bw() +
  theme( axis.text = element_text( size = 14 ),
         axis.title = element_text( size = 16) ) +
  labs( x = "3PA", y = "3P%")

# No apparent relationship between 3P% and 3PA.
```

```{r, echo = FALSE}
# Fit team data to beta distribution and extract coefficients
team.fit <- MASS::fitdistr( x = team.three$`3P%`, densfun = "beta",
                            start = list( shape1 = 100, shape2 = 100 ),
                            method = "Nelder-Mead")

team.coefs <- coefficients( team.fit )
```

```{r, warning = FALSE, echo = FALSE, eval = FALSE}
# Visualize fit

ggplot( team.three, aes( x = `3P%` ) ) +
  geom_histogram( aes( y =  after_stat( density ) ), 
                  bins = 15, color = "black", fill = "gray75" ) +
  stat_function( fun = dbeta,
                 args = list( shape1 = team.coefs[1],
                              shape2 = team.coefs[2] ),
                 col = "red",
                 size = 1 ) +
  theme_bw() +
  theme( axis.text = element_text( size = 14 ),
         axis.title = element_text( size = 16) ) +
  scale_x_continuous( limits = c( 0.3, 0.425 ) ) +
  labs( x = "3P%", y = "Density" )
```

```{r, echo = FALSE}
# Use fitted beta parameters to shrink Boston and Miami team data for 2023 reg season and playoffs

bm.stats <- bm.stats |>
  dplyr::mutate( post.alpha = team.coefs[1] + `3P`,
                 post.beta = team.coefs[2] + `3PA` - `3P`,
                 `adj3P%` = post.alpha / ( post.alpha + post.beta ) ) |>
  dplyr::arrange( Tm, time )

plot.df <- bm.stats |>
  dplyr::select( Tm, `3P%`, `adj3P%`, time ) |>
  tidyr::pivot_longer( cols = c( `3P%`, `adj3P%` ), values_to = "3P%", names_to = "value" ) |>
  dplyr::mutate( value = case_when( value == "3P%" ~ "Observed",
                                    value == "adj3P%" ~ "Adjusted" ) )
```

```{r, echo = FALSE}
# bm.stats$time <- factor( bm.stats$time, levels = c( "Season", "Playoffs" ) )

plot.df$value <- factor( plot.df$value, levels = c( "Observed", "Adjusted" ) )

ggplot( plot.df |> dplyr::filter( time == "Playoffs"), 
        aes( x = `3P%`, y = Tm, fill = value ) ) +
  geom_segment( data = bm.stats |> dplyr::filter( time == "Playoffs"),
                aes(x = `3P%`, xend = `adj3P%`,
                    y = Tm, yend = Tm ),
                size = 1,
                inherit.aes = FALSE ) +
  geom_point( shape = 21, size = 8, stroke = 1 ) +
  scale_fill_manual(values = c( "darkblue", "darkgreen" ),
                    guide = guide_legend( override.aes = list(size = 7) ) ) +
  theme_bw() +
  theme( legend.position = "top",
         legend.title = element_blank(),
         legend.text = element_text(size=12),
         legend.justification="center",
         axis.text = element_text( size = 14 ),
         axis.title = element_text( size = 16),
         axis.text.y = element_text( angle = 90, hjust = 0.5 ) ) +
  scale_y_discrete( labels = c( "Boston", "Miami" ) ) +
  scale_x_continuous( limits = c( 0.30, 0.45 ) ) +
  labs( x = "3P%", y = "Team")
```

```{r}
season.dist <- apply( X = bm.stats,
                      MARGIN = 1,
                      FUN = function(x){
                       alpha <- as.numeric( x[6] )
                       beta <- as.numeric( x[7] )
                       rbeta( 1E6, shape1 = alpha, shape2 = beta )
                      })
```









```{r}
fit <- gamlss( cbind( `3P`, `3PA` - `3P` ) ~ log( `3PA` ) ,
               data = season.three,
               family = gamlss.dist::BB( mu.link = "identity", sigma.link = "log" ) )

plot( fit )
```



```{r}
season.three <- season.three |>
  dplyr::mutate( mu = fitted( fit, parameter = "mu" ),
                 sigma = fitted( fit, parameter = "sigma" ),
                 alpha0 = mu / sigma,
                 beta0 = (1 - mu) / sigma,
                 alphai  = alpha0 + `3P`,
                 betai =  beta0 + `3PA` - `3P`,
                 eb = alphai / ( alphai + betai ) )
```

```{r}
ggplot( three.dat, aes( x = `3P%`, y = eb, fill = `3PA` ) ) +
  geom_point(size = 1, shape = 21, color = "#00000000") + 
  scale_fill_gradient("3P Attempts",
                      guide = guide_colorbar(title.position = "top",
                                             nrow = 1,
                                             title.hjust=0.5,
                                             ticks.colour="black",
                                             ticks.linewidth=1,
                                             frame.colour="black",
                                             frame.linewidth=0.5)) +
  theme_bw() +
  theme( legend.position="top",
         legend.key.size = unit(0.3, "cm"),
         legend.key.width = unit(1,"cm"),
         legend.text = element_text(size=10),
         legend.title=element_text(size=12),
         legend.box.margin=margin(-10,-10,-10,-10),
         legend.justification="center",
         axis.text = element_text( size = 14 ),
         axis.title = element_text( size = 16) ) +
  labs( x = "3P%", y = "Adjusted 3P%")
```


```{r}
library( ggpointdensity )

ggplot( data = season.three, aes( x = `3PA`, y = `3P%` ) ) +
  geom_hline( yintercept = mean( season.three$`3P%` ), color = "black", linewidth = 0.5 ) + 
  geom_pointdensity( adjust = 2, size = 1 ) +
  # geom_point( aes( x = `3PA`, y = eb ), color = "purple", alpha = 0.5 ) +
  scale_color_gradient(guide = guide_colorbar(title = "Nearest Neighbors",
                                              title.position = "top",
                                              nrow = 1,
                                              title.hjust=0.5,
                                              ticks.colour="black",
                                              ticks.linewidth=1,
                                              frame.colour="black",
                                              frame.linewidth=0.5 ) ) +
  geom_line( aes(x = `3PA`, y = `mu` ), color = "red", size = 1 ) +
  theme_bw() +
  theme( legend.position="top",
         legend.key.size = unit(0.3, "cm"),
         legend.key.width = unit(1,"cm"),
         legend.text = element_text(size=10),
         legend.title=element_text(size=12),
         legend.box.margin=margin(-10,-10,-10,-10),
         legend.justification="center",
         axis.text = element_text( size = 14 ),
         axis.title = element_text( size = 16) ) +
  labs( x = "3P Attempts", y = "3P%")
```

```{r}
year <- "2023"
playoff.url <- paste0( "https://www.basketball-reference.com/playoffs/NBA_", year, "_totals.html" )
playoff.page <- read_html( playoff.url )

playoff.cols <- playoff.page %>% 
  html_nodes("table#totals_stats > thead > tr > th") %>%
  html_text()

playoff.cols <- playoff.cols[ -c(1) ]

playoff.dat <- data.frame( 
  t( matrix( data = playoff.page %>% 
               html_nodes("table#totals_stats > tbody > tr > td") %>%
               html_text(),
             nrow = 29 )
     ) 
  )

colnames( playoff.dat ) <- playoff.cols

playoff.dat <- playoff.dat |>
  dplyr::mutate( Age = as.numeric( Age ),
                 across( c(5:29), as.numeric ) ) |>
  tibble::add_column( Season = 2023, .after = 3 )
```

```{r}
playoff.threes <- playoff.dat |>
  dplyr::select( Player, Season, Tm, MP, `3P`, `3PA`, `3P%` ) |>
  tidyr::drop_na()

coefs <- tidy(fit)$estimate

playoff.threes <- playoff.threes |>
  dplyr::mutate( mu = coefs[1] + coefs[2] * log( `3PA` ),
                 sigma = exp( coefs[3] ),
                 alpha0 = mu / sigma,
                 beta0 = (1 - mu) / sigma,
                 alphai  = alpha0 + `3P`,
                 betai =  beta0 + `3PA` - `3P`,
                 eb = alphai / ( alphai + betai ) )
```

```{r}
poi <- c( "Gabe Vincent", "Max Strus", "Duncan Robinson", "Caleb Martin" )

heat.reg <- three.dat |>
  dplyr::filter( Player %in% poi, Season == 2023 ) |>
  dplyr::mutate( type = "Regular Season")

heat.post <- playoff.threes |>
  dplyr::filter( Player %in% poi ) |>
  dplyr::mutate( type = "Playoffs" )

poi.dat <- rbind( heat.reg, heat.post) |>
  dplyr::arrange( Player )

heat.reg.sum <- three.dat |>
  dplyr::filter( Player %in% poi, Season == 2023 ) |>
  dplyr::summarise( total.3P = sum( `3P` ),
                    total.3PA = sum( `3PA` ) ) |>
  dplyr::mutate( total.perc = total.3P / total.3PA,
                 type = "Regular Season")

heat.playoff.sum <- playoff.threes |>
  dplyr::filter( Player %in% poi ) |>
  dplyr::summarise( total.3P = sum( `3P` ),
                    total.3PA = sum( `3PA` ) ) |>
  dplyr::mutate( total.perc = total.3P / total.3PA,
                 type = "Playoffs")
```

```{r}
playoff.threes |>
  dplyr::filter( Player %in% c( "Jayson Tatum", "Jaylen Brown", "Al Horford", "Derrick White") ) |>
  dplyr::mutate( type = "Playoffs" )

three.dat |>
  dplyr::filter( Player %in% c( "Jayson Tatum", "Jaylen Brown", "Al Horford", "Derrick White"),
                 Season == 2023 ) |>
  dplyr::mutate( type = "Regular Season" )
```

```{r}
mia.url <- "https://www.basketball-reference.com/playoffs/2023-nba-eastern-conference-finals-heat-vs-celtics.html"

mia.page <- read_html( mia.url )

mia.cols <- mia.page %>% 
  html_nodes("table#MIA > thead > tr > th") %>%
  html_text()

mia.cols <- mia.cols[ -c(1:5) ]

mia.dat <- data.frame( 
  t( matrix( data = mia.page %>% 
               html_nodes("table#MIA > tbody > tr > td") %>%
               html_text(),
             nrow = 29 )
     ) 
  )

colnames( mia.dat ) <- mia.cols

mia.dat <- mia.dat[ , 1:20 ]

mia.dat <- mia.dat |>
  dplyr::mutate( across( 2:20, as.numeric ) ) |>
  tibble::add_column( Season = 2023, .after = 3 )
```
















